---
title: "Trend Surface Analysis with R (part I)"
output: html_notebook
---

This Notebook (the first-of-four) serves to introduce a user to a series of exercises focused on higher order Trend Surface analysis with R.

It consists of two sections:

1. Explore the dataset
2. Deterministic Interpolation Techniques
	a) Voronoi
	b) Inverse Distance Weighting

$\color{red}{\text{\textbf{REQUIRED}}}$: 
You are required to insert your outputs and any comment into this document. The document you submit should therefore contain the existing text in addition to:

- Plots and other outputs from executing the code chunks
- Discussion of your plots and other outputs as well as conclusions reached.
- This should also include any hypotheses and assumptions made as well as factors that may affect your conclusions.

```{r install }
options(prompt="> ", continue="+ ", digits=3, width=70, show.signif.stars=F, repr.plot.width=7, repr.plot.height=7)
rm(list=ls())

# Install necessary packages: You only need to run this part once
install.packages(c("sf", "gstat", "stars"))

library(sf)    # 'simple features' representations of spatial objects
library(gstat) # geostatistics
library(stars) # gridded data structures ("rasters")
```
## 1. Explore the dataset

```{r read-dataset }
#-- read
file = 'cptFlatsAquifer_watertable4.txt'
```

```{r import }
#-- import
cfaq <- read.csv(file, header = 1, sep = ',', dec = '.')
```

```{r look }
#-- look
head(cfaq ,3)
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- What is the purpose of producing a map of the the elevation of the top of the aquifer over the study area? In other words, who would use the map and for what purpose?

```{ r ans-01}
[click in this cell and type your answer here]
```

```{r str-cfaq }
str(cfaq)
```

```{r crs-and-look }
#- set as a spatial feature with xy coords an existing projection
cfaq.sf <- st_as_sf(cfaq, coords=c("long", "lat"), crs = 4326) #wgs84
#str(cfaq)
st_crs(cfaq.sf)
```

```{r transform-crs }
#- transform to local crs
cfaq.sf <- st_transform(cfaq.sf, crs = 32734) #utm 34s
cfaq.sf$X <- st_coordinates(cfaq.sf)[, "X"]
cfaq.sf$Y <- st_coordinates(cfaq.sf)[, "Y"]
```

```{r str-sf }
str(st_coordinates(cfaq.sf))
```

```{r summarize-coords }
#- summarize the coordinates
summary(st_coordinates(cfaq.sf), digits = 3)
```

```{r dim }
dim(cfaq.sf)
```

```{r summarize-cfaq }
#- summarize the data
summary(cfaq.sf)
```

```{r bbox-cfaq-sf) }
#- bounding box
st_bbox(cfaq.sf)
```

```{r range-cfaq-sf }
#- a couple of range measurments
range(cfaq.sf$waterLevel); diff(range(cfaq.sf$waterLevel))
range(cfaq.sf$elevation); diff(range(cfaq.sf$elevation))
range(cfaq.sf$depth); diff(range(cfaq.sf$depth))
```

$\color{red}{\text{\textbf{QUESTION}}}$: 

- How many observations are there? What was recorded at each point?

```{ r ans-02}
[click in this cell and type your answer here]
```

- What are the geographic limits of the study area? What is its area, in km2?

```{ r ans-03}
[click in this cell and type your answer here]
```
```{r plot-names }
plot(st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
     #plot(st_coordinates(cfaq.sf)[,1] ~ st_coordinates(cfaq.sf)[,2],
     pch=20, cex=0.4, col="blue", asp=1,
     xlab="Lo 19 E", ylab="Lo 19 N")
grid()
text(st_coordinates(cfaq.sf)[,1], st_coordinates(cfaq.sf)[,2],
     #st_coordinates(cfaq.sf)[,2], st_coordinates(cfaq.sf)[,1],
     #round(cfaq$waterLevel), adj=c(0.5,0.5))
     cfaq$name, adj=c(0.5,0.5))
#text(cfaqN, round(aq$z), adj=c(0.5,0.5))
title("Elevation of aquifer, m")
```
```{r plot-elevation }
plot(#st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
  st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
  cex=0.8*cfaq$water_level/max(cfaq$waterLevel),
  col="blue", bg="red", pch=21, asp=1,
  xlab="Lo19 E", ylab="Lo19 N")
grid()
title("Elevation of aquifer, m")
```

```{r plot-elevation-viridis }
plot(#st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
  st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
  pch=21,
  xlab="UTM19_E", ylab="UTM19_N",
  bg=sp::bpy.colors(length(cfaq$waterLevel))[rank(cfaq$waterLevel)],
  cex=0.9)#*cfaq$waterLevel/max(cfaq$waterLevel), asp=1)
grid()
title("Elevation of aquifer, m")
```

$\color{red}{\text{\textbf{QUESTION}}}$: 

- Describe the spatial pattern of the elevations. Do nearby points have similar values? Is there a trend across the whole area? Are there local exceptions to the trend?

```{ r ans-02}
[click in this cell and type your answer here]
```

## 2. Deterministic Interpolation Techniques

Before we delve into higher order interpolation techniques; let us first explore two extremly useful methods we can use to model a continuous variable (like an elevation surface or meterological observations).

a) Thiessen Polygons
b) Inverse Distance Weighting (IDW)

$\color{red}{\text{\textbf{REMEMBER:}}}$: our goal is to estimate values where non exist. And depending on the application (and the data we have access to) higher order methods might not be necessary!

### a) Thiessen Polygons

The oldest method is simply to divide the area of interest based on proximity to sample points. The result is a tesselated surface of Thiessen polygons (also called a Voronoi Diagram)

```{r voronoi }
# Voronoi tesselation
pnts <- st_union(cfaq.sf)
voronoi_grid <- st_voronoi(pnts)
```

```{r plot-voronoi }
#-- plot
plot(voronoi_grid, col = NA)
points(st_coordinates(cfaq.sf)[,1], st_coordinates(cfaq.sf)[,2],
       pch=21,
       bg=sp::bpy.colors(length(cfaq$waterLevel))[rank(cfaq$waterLevel)],
       cex=0.4*cfaq$waterLevel/max(cfaq$waterLevel))
grid()
title("Elevation of aquifer, m.a.s.l.")
```

### b) Inverse Distance Weighting (IDW)

An extremely popular interpolation method calculates an average value for estimated locations using values from nearby weighted locations.

```{r range-cfaq-coords }
range(st_coordinates(cfaq.sf)[,2]); range(st_coordinates(cfaq.sf)[,1])
```

Create a 500-m grid

```{r create-grid }
#- create grid
cfaq.bb <- st_bbox(cfaq.sf)
cfaq.bb[c("xmin","ymin")] <-   floor(cfaq.bb[c("xmin","ymin")]/1000)*1000
cfaq.bb[c("xmax","ymax")] <- ceiling(cfaq.bb[c("xmax","ymax")]/1000)*1000

#-- as a stars object
grid <- st_as_stars(cfaq.bb, dx = 500)
grid
```

```{r dim-grid }
dim(grid)
```

```{r summarize-grid }
summary(grid)
```

```{r bbox(-grid }
st_bbox(grid)
```

```{r plot-grid }
plot(grid); grid()
```

**IDW interpolation with power at 0.5, 1, 2.5 and 10** and plot the result

```{r idw-power }
#- idw
idw_results <- list()
idp_values <- c(0.5, 1, 2.5, 10)
```

```{r idw-loop}
#- idw
for (idp in idp_values) {
  idw_result <- idw(log(waterLevel) ~ 1, cfaq.sf, grid, idp = idp)
  #idw <- gridkm
  #idw <- idw_result
  idw_results[[as.character(idp)]] <- idw_result
}
```

```{r plot-idw }
# Plot IDW results with overlay of original points
num_plots <- length(idp_values)
par(mfrow = c(2, 2))  # Arrange plots in 2 rows and 2 columns
# Set the size of the individual plots
plot_width <- 9
plot_height <- 9

# Set the margin and space between plots
margin_size <- 0.5
space_between_plots <- 0.3

# Calculate the total plot area
total_width <- plot_width * 2 + space_between_plots
total_height <- plot_height * 2 + space_between_plots

# Set par settings for plot layout
par(mfrow = c(2, 2), mar = c(0.6, 0.6, 0.8, 0.1),#margin_size, margin_size, margin_size, margin_size),
    oma = c(0, 0, 0, 0), mgp = c(0.5, 0.7, 0))

min.x <- floor(min(cfaq.sf$X)/1000)*1000
max.x <- ceiling(max(cfaq.sf$X)/1000)*1000

min.y <- floor(min(cfaq.sf$Y)/1000)*1000
max.y <- ceiling(max(cfaq.sf$Y)/1000)*1000

for (i in 1:num_plots) {
  idp <- idp_values[i]
  idw_result <- idw_results[[as.character(idp)]]
  title <- paste("IDW @ (IDP =", idp, ")", sep = " ")

  # Plot interpolated surface
  image(idw_result, main = title, col = rainbow(100),
        xlim = c(min.x, max.x), ylim = c(min.y, max.y),
        xlab = "", ylab = "")

  # Overlay original points
  points(cfaq.sf$X, cfaq.sf$Y, pch = 21,
         bg = adjustcolor("black", alpha.f = 0.2),
         col = "black",
         cex = 0.9, lwd = 0.5)

  contour(idw_result, add = TRUE, nlevels = 10, col = "black")
}

# Reset layout to default
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

$\color{red}{\text{\textbf{QUESTION}}}$: 

- Why do we need interpolation methods that go deeper / further than the two presented here?

```{ r ans-03}
[click in this cell and type your answer here]
```

- Comment on the quality of the IDW
$\color{blue}{\text{\textbf{Hint}}}$: Think about the subjectivity of the power function and what this means. Also consider what happens when we overlay the voronoi with the various idw interpolations. What does this mean?

```{r plot-idw-voronoi }
# Plot IDW results with overlay of original points
num_plots <- length(idp_values)
par(mfrow = c(2, 2))  # Arrange plots in 2 rows and 2 columns
# Set the size of the individual plots
plot_width <- 9
plot_height <- 9

# Set the margin and space between plots
margin_size <- 0.5
space_between_plots <- 0.3

# Calculate the total plot area
total_width <- plot_width * 2 + space_between_plots
total_height <- plot_height * 2 + space_between_plots

# Set par settings for plot layout
par(mfrow = c(2, 2), mar = c(0.6, 0.6, 0.8, 0.1),#margin_size, margin_size, margin_size, margin_size),
    oma = c(0, 0, 0, 0), mgp = c(0.5, 0.7, 0))

for (i in 1:num_plots) {
  idp <- idp_values[i]
  idw_result <- idw_results[[as.character(idp)]]
  title <- paste("IDW @ (IDP =", idp, ")", sep = " ")

  # Plot interpolated surface
  image(idw_result, main = title, col = rainbow(100),
        xlim = c(min.x, max.x), ylim = c(min.y, max.y),
        xlab = "", ylab = "")

  # Overlay Voronoi polygons
  plot(voronoi_grid, add = TRUE, border = "black", col = NA, type="l", lty=2) #lwd = 1,

  # Overlay original points
  points(cfaq.sf$X, cfaq.sf$Y, pch = 21,
         bg = adjustcolor("black", alpha.f = 0.2),
         col = "black",
         cex = 0.9, lwd = 0.5)
}

# Reset layout to default
par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))
```

```{ r ans-04}
[click in this cell and type your answer here]
```
- How can the quality of the IDW improve?

```{ r ans-05}
[click in this cell and type your answer here]
```
- How can the quality of the IDW improve?
$\color{blue}{\text{\textbf{Hint}}}$: Consider the location and density of the sample (input) points

```{ r ans-05}
[click in this cell and type your answer here]
```