---
title: "Trend Surface Analysis with R (part III)"
output: html_notebook
---
##-- arkriger: August 2023

In this third Notebook on *higher order* **_Trend Surface analysis with R_** we introduce;

1. General Additive Models (GAM) and
2. Thin plate splines

$\color{red}{\text{\textbf{REQUIRED}}}$: 
**You are required to insert your outputs and any comment into this document.** The document you submit should therefore contain the existing text in addition to:

- Plots and other outputs from executing the code chunks
- Discussion of your plots and other outputs as well as conclusions reached.
- This should also include any hypotheses and assumptions made as well as factors that may affect your conclusions.

## 1. General Additive Models

We now truly start our journey into higher order trend surface analysis. GAMs are very similar to multiple `linear regression`; where each predictor  typically has a smooth function. The advantage is that GAMs can capture intricate and complex patterns traditional `linear models` often miss.

The disadvantage of these methods are their computational complexity and the risk of overfitting. It is pretty dangerous to extrapolate outside the range of calibration (i.e.: beyond the observation points).

```{r install }
options(prompt="> ", continue="+ ", digits=3, width=70,  show.signif.stars=F, repr.plot.width=7, repr.plot.height=7)
rm(list=ls())

# Install necessary packages: You only need to run this part once
##- install.packages(c("sf", "gstat", "ggplot2", "gridExtra","units", "terra","mgcv","fields"))

#library(repr)
library(sf) # 'simple features' representations of spatial objects
library(gstat) # geostatistics
library(ggplot2) # geostatistics
library(gridExtra)
library(units) # units of measure
library(terra) # gridded data structures ("rasters")
library(mgcv)
library(fields)
```

We've already covered creating a 500m grid and 2nd-order prediction and interpolation with previous Notebooks and move through the introduction quickly.

```{r read-dataset }
#-- read
file = 'cptFlatsAquifer_watertable4.txt'
```

```{r import-grid-2nd-ols }
#-- import
cfaq <- read.csv(file, header = 1, sep = ',', dec = '.') #sep = '\t',
#- set crs
cfaq.sf <- st_as_sf(cfaq, coords=c("long", "lat"), crs = 4326) #wgs84
#- transform to local crs
cfaq.sf <- st_transform(cfaq.sf, crs = 32734) #utm 34s
cfaq$X <- st_coordinates(cfaq.sf)[, "X"]
cfaq$Y <- st_coordinates(cfaq.sf)[, "Y"]

model.ts2 <- lm(waterLevel ~ X + Y + I(X^2) + I(Y^2) + I(X*Y), data=drop_units(cfaq))

#--  create grid
#(n.col <- length(seq.e <- seq(min.x <- floor(min(st_coordinates(cfaq.sf)[,1])/1000)*1000,
#                              max.x <- ceiling(max(st_coordinates(cfaq.sf)[,1])/1000)*1000, by=1000)))
#(n.col <- length(seq.e <- seq(min.x <- floor(min(st_coordinates(cfaq.sf)[,2])/1000)*1000,
#                              max.x <- ceiling(max(st_coordinates(cfaq.sf)[,2])/1000)*1000, by=1000)))
(n.col <- length(seq.e <- seq(min.x <- floor(min(cfaq$X)/1000)*1000,
                              max.x <- ceiling(max(cfaq$X)/1000)*1000, by=1000)))
(n.row <- length(seq.n <- seq(min.y <- floor(min(cfaq$Y)/1000)*1000,
                              max.y <- ceiling(max(cfaq$Y)/1000)*1000, by=1000)))

#we want a XXXXm grid
grid <- rast(nrows = n.row, ncols = n.col,
             xmin=min.x, xmax=max.x,
             ymin=min.y, ymax=max.y, crs = st_crs(cfaq.sf)$proj4string,
             resolution = 500, names="waterLevel")

values(grid) <- NA_real_

grid.df <- as.data.frame(grid, xy = TRUE, na.rm = FALSE)
names(grid.df)[1:2] <- c("X", "Y") # match the names of the point dataset
summary(grid.df)

#--
pred.ts2 <- predict.lm(model.ts2,
                       newdata = grid.df,
                       interval = "prediction", level = 0.95)

#-- add the three prediction fields (fit, lower, upper) to the data
grid.df[, 3:5] <- pred.ts2
names(grid.df)[3:5] <- c("ts2.fit", "ts2.lwr", "ts2.upr")
```

```{r look }
#-- look
head(cfaq ,3)
```

We start through first exploring whether a curve _(a smoothing function)_ better captures the relationships between location (x, y) and the aquifer water level. We do this through the local polynomial regression via the `leoss` function built into `ggplot2` as `geom_smooth`.

```{r plot-ggplot2-leoss }
##- plot
g1 <- ggplot(drop_units(cfaq), aes(x=Y, y=waterLevel)) +
  geom_point() +
  geom_smooth(method="loess") +
  labs(y = "elevation [m]")
g2 <- ggplot(drop_units(cfaq), aes(x=X, y=waterLevel)) +
  geom_point() +
  geom_smooth(method="loess") +
  labs(y = "elevation [m]")
grid.arrange(g1, g2, ncol = 2)
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- Do these marginal relations appear to be linear in the predictors?

```{ r ans-01}
[click in this cell and type your answer here]
```

**Fitting the GAM**

We work with the **Mixed GAM Computation Vehicle** `mgcv` library default smoothing setting and explore **thin plate splines**, as an alternate smoothing function later.

```{r fit-gam-summary}
model.gam <- gam(waterLevel ~ s(X, Y, k = 29), data=drop_units(cfaq))
summary(model.gam)
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- How well does this model fit the calibration observations?

```{ r ans-02}
[click in this cell and type your answer here]
```

**Since we have a 2nd-order trend surface; lets compare.**

we do this with numbers and graphs.

```{r residuals-gam-summary }
#- residuals
resid.gam <- residuals(model.gam)
summary(resid.gam)
```

```{r residuals-2nd-ols-summary}
summary(residuals(model.ts2))
```
```{r hist-gam-2nd-ols-residuals }
#regular histogram
par(mfrow=c(1,2))
hist(resid.gam)#, xlim=c(-5, 5))#,
     #breaks=c(min(cfaq$waterLevel), max(cfaq$waterLevel), by=2)), main="Residuals from GAM")
rug(residuals(model.gam))
hist(residuals(model.ts2))#, xlim=c(-30, 30))#,
     #breaks=c(min(cfaq$waterLevel), max(cfaq$waterLevel), by=2), main="Residuals from 2nd-order OLS trend")
rug(residuals(model.ts2))
par(mfrow=c(1,1))
```

```{r ggplot2-hist-gam-2nd-ols-residuals }
#-- ggplot2
g1 <- ggplot(data=as.data.frame(resid.gam), aes(resid.gam)) +
  geom_histogram(#breaks=seq(-20,20,by=2),
                 fill="lightblue", color="black", alpha=0.9) +
  geom_rug() +
  labs(title = "Residuals from GAM",
       x = expression(paste(Delta, m)))
g2 <- ggplot(data=as.data.frame(residuals(model.ts2)), aes(residuals(model.ts2))) +
  geom_histogram(#breaks=seq(-20,20,by=2),
                 fill="lightgreen", color="darkblue", alpha=0.9) +
  geom_rug() +
  labs(title = "Residuals from 2nd order polynomial trend surface",
       x = expression(paste(Delta, m)))
grid.arrange(g1, g2, nrow=1)
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- Which histogram shows the narrowest spread?

```{ r ans-03}
[click in this cell and type your answer here]
```

```{r plot-residuals-gam }
#-- residuals as a bubble plot. show the size of the residuals by the size of a point, and the polarity (positive vs. negative)
cfaq$resid.gam <- resid.gam
ggplot(data = drop_units(cfaq)) +
  aes(x=X, y=Y, size = abs(resid.gam),
      col = ifelse((resid.gam < 0), "red", "green")) +
  geom_point(alpha=0.7) +
  scale_size_continuous(name = expression(paste(plain("residual ["),
                                                reDelta, m, plain("]"))),
                        breaks=seq(0,12, by=2)) +
  scale_color_manual(name = "polarity",
                     labels = c("negative","positive"),
                     values = c("red","green","blue"))
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- Does there appear to be any local spatial correlation of the residuals?

```{ r ans-04}
[click in this cell and type your answer here]
```

```{r plot-gam-3D}
#- gam as 3D surface
plot.gam(model.gam, rug = FALSE, se = FALSE, select=1,
         scheme=1, theta=110+130, phi=30)
```

```{r plot-gam-3D-std-error }
#-3D with 1 standard error of fit
vis.gam(model.gam, plot.type="persp", color="terrain",
        theta=160, zlab="elevation", se=1.96)
```

**GAM `RMSE`and predict the model onto our 500m grid**

```{r rmse }
#-- rmse
(rmse.gam <- sqrt(sum(residuals(model.gam)^2)/length(residuals(model.gam))))
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- Does there appear to be any local spatial correlation of the residuals?

```{ r ans-05}
[click in this cell and type your answer here]
```

```{r predict-gam-summary }
#-- predict aquifer elevation, standard error of prediction using the fitted GAM
tmp <- predict.gam(object=model.gam,
                   newdata=grid.df,
                   se.fit=TRUE)
summary(tmp$fit)
```

```{r gam-std-error-summary }
summary(tmp$se.fit)
```

```{r str }
str(tmp)
```

```{r add-gam-to-griddf}
#-- add these to the data.frame of the spatial grid with the as.numeric(1d array to vector)
grid.df$pred.gam <- as.numeric(tmp$fit)
grid.df$pred.gam.se <- as.numeric(tmp$se.fit)
```
```{r plot-gam }
#-- plot
grid.gam <- grid
values(grid.gam) <- grid.df$pred.gam
plot(grid.gam, col = rainbow(100), main="GAM prediction"); grid()
points(st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
#       pch=16,
       col=ifelse(cfaq$resid.gam < 0, "red", "green"),
       cex=2*abs(cfaq$resid.gam)/max(abs(cfaq$resid.gam)),
                                     pch=21,
                                     bg = adjustcolor("black", alpha.f = 0.1),
                                     #col = "black",
                                     #cex = 0.9,
                                     lwd = 0.5)
```

$\color{red}{\text{\textbf{QUESTION}}}$: 
- How well does the GAM trend surface fit the points? Are there obvious problems?

```{ r ans-06}
[click in this cell and type your answer here]
```
```{r plot-gam-uncertainty }
#- standard errors of prediction
grid.gam.se <- grid
values(grid.gam.se) <- grid.df$pred.gam.se
plot(grid.gam.se, main="GAM prediction standard error",
     col=cm.colors(64))
grid()
points(st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1], pch=16,
       col="grey")
```

We see our GAM standard errors follow a similar pattern to the polynomial surfaces with concentrations higher along the edges but what we really want to know is where the GAM surface differs from the traditional parametric

```{r plot-diff-gam-2nd-ols}
summary(grid.df$diff.gam.ols <- grid.df$pred.gam - grid.df$ts2.fit)

grid.diff.gam.ols <- grid
values(grid.diff.gam.ols) <- grid.df$diff.gam.ols
plot(grid.diff.gam.ols,
     main="Difference (GAM - 2nd order trend surface) predictions, [m]",
     col=topo.colors(64))
grid()
```

Here the positive differences are where the GAM predicts higher values than the trend surface.

$\color{red}{\text{\textbf{QUESTION}}}$: 
- Where are the largest differences between the GAM and 2nd order OLS trend surface predictions? Explain why, considering how the two surfaces are computed.

```{ r ans-07}
[click in this cell and type your answer here]
```

## 2. Thin-plate spline interpolation

Thin-plate splines are equivalent to a _thin and flexible_ plate warped to fit a dataset. We can do this with a single _plane_ surface similar to a 1st-degree polynomial all the way through to an extremely supple surface that perfectly fit every observation.

Generally we choose a balanced approach because overfitting introduces noise.

```{r setup-tps }
#-- set up thin plate spline
cfaq.tps <- cfaq[, c("X","Y", "waterLevel")]
cfaq.tps$coords <- matrix(c(cfaq.tps$X, cfaq.tps$Y), byrow=F, ncol=2)
str(cfaq.tps$coords)
```

```{r tps-fields }
surf.1 <- fields::Tps(cfaq.tps$coords, cfaq.tps$waterLevel)
```

```{r tps-summary}
summary(surf.1)
```

```{r tps-grid }
#-- predict over the study area
grid.coords.m <- as.matrix(grid.df[, c("X", "Y")], ncol=2)
str(grid.coords.m)
```

```{r predict-tps }
surf.1.pred <- predict.Krig(surf.1, grid.coords.m)
summary(grid.df$pred.tps <- as.numeric(surf.1.pred))
```

```{r tps-residuals-summarize }
#-- compute and summarize residuals
grid.tps <- grid
values(grid.tps) <- surf.1.pred
tmp <- extract(grid.tps, st_coordinates(cfaq.sf))
#names(tmp)

summary(cfaq.sf$resid.tps <- (cfaq$waterLevel - tmp$waterLevel))
```

```{r hist-tps-residuals }
#- histogram of the residuals
hist(cfaq.sf$resid.tps, main="Thin-plate spline residuals", breaks=16)
rug(cfaq.sf$resid.tps)
```

```{r plot-tps }
#- plot
plot(grid.tps, col = rainbow(100),
     main = "2D Thin-plate spline surface")
points(st_coordinates(cfaq.sf)[,2] ~ st_coordinates(cfaq.sf)[,1],
       #       pch=16,
       col=ifelse(cfaq.sf$resid.tps < 0, "red", "green"), #"black", "grey"),
       cex=2*abs(cfaq.sf$resid.tps)/max(abs(cfaq.sf$resid.tps)),
       pch=21,
       bg = adjustcolor("black", alpha.f = 0.1),
       #col = "black",
       #cex = 0.9,
       lwd = 0.5)
```

As with the other interpolation plots; the red and green illustrate positive and negative errors with size representative of magnitude. Notice how closely the surface fits the points.

Similarly; what we really want to know is where the thin-plate spline differs from the GAM surface.

```{r plot-diff-tps-gam }
#- show the surface
grid.diff.tps.gam <- grid
grid.diff.tps.gam <- grid.tps - grid.gam
values(grid.diff.tps.gam) <- (values(grid.tps) - values(grid.gam))
plot(grid.diff.tps.gam,
     col = topo.colors(64),
     main = "Thin-plate spline less GAM fits, [m]",
     xlab = "X", ylab = "Y")
```

$\color{green}{\text{\textbf{A BREATHER}}}$: 

Before we continue onto the next Notebook and Generalized Least Squares; we might want to take a break and possibly understand why we would want to go further than this pretty good surface!

One of the principles of Spatial Data Science is; **data is spatially autocorrelated**. Patterns and particularly residuals (errors) have a spatial structure / are connected.

In other words; **_'everything is related to everything else, but near things are more related than distant things'_** - [The First Law of Geography, according to Waldo Tobler](https://en.wikipedia.org/wiki/Tobler%27s_first_law_of_geography#cite_note-1)

We can model the spatial structure of residuals with a **variogram**; and then use this model to refine a trend surface with Generalized Least Squares.

Recall our formula for a 1st- (with more degrees of freedom for the  2nd-) order surface;

$$
    Z = ðœ·_0 + ðœ·_1x + ðœ·_2y + Ïµ   
$$

In effect we do away with the broad $Ïµ$ term _---applied to the entire study area---_ and introduce local terms to mininize the overall error

```{r emphirical-variogram}
#- emphirical variogram
cfaq$fit.ts2 <- fitted(model.ts2)
cfaq.sf$fit.ts2 <- fitted(model.ts2)
cfaq$res.ts2 <- residuals(model.ts2)
cfaq.sf$res.ts2 <- residuals(model.ts2)
```

```{r plot-variogram-fit}
vr.c <- variogram(res.ts2 ~ 1, loc = cfaq.sf, cutoff = 50000, cloud = T)
vr <- variogram(res.ts2 ~ 1, loc = cfaq.sf, cutoff = 50000)
p1 <- plot(vr.c, col = "blue", pch = 20, cex = 0.5,
           xlab = "separation [m]", ylab = "semivariance [m^2]")
p2 <- plot(vr, plot.numbers = T, col = "blue", pch = 20, cex = 1.5,
           xlab = "separation [m]", ylab = "semivariance [m^2]")
print(p1, split = c(1,1,2,1), more = T)
print(p2, split = c(2,1,2,1), more = F)
```

<div class="alert alert-block alert-info"><b> </b> The emphrical variogram is characterised by three parameters. Here we scale the values by looking at the graph. Later we precisely model them. </div>

- $c_0$ **the nugget**: the semivariance at zero.  

- $c$ **the sill**: the maximum variance when point-pairs are widely spaced.

- $a$ **the range**: the seperation where there is no more spatial autocorrelation. i.e.: where the semivariance reaches **the sill**.

$\color{red}{\text{\textbf{QUESTION}}}$: 
- What are the estimated sill, range, and nugget of this variogram?

```{ r ans-07}
[click in this cell and type your answer here]
```

